---
title: "Efficient Data Handling"
author: "Michal Kubišta"
date: "2018/03/21"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(magrittr)
library(data.table)
library(tidyverse)
library(xaringan)
```

class: center, middle

# <a href="https://www.keylink.net.au/etl">What's ETL?</a>
<img src = www/etl.jpg>

---
class: center, middle

# Efficiency?

---

# Terminology
- data.frame / data.table
- vector
- list

---

# Packages
- software extensions
- on 2017-01 there were 10 000 official ones
    - fraction of the real number
- today: two framework packages
    - tidyverse
    - data.table
- framework: offer a wide range of commands to cover the majority of common functionalities
- new functions or changing the old ones
    - base::merge() **VS** dplyr::inner_join()
    - base::merge() **VS** data.table::merge()

---

# Key words - single table
- read
- filter
- select
- arrange
- mutate
- group by & summarize
- spread & gather
- write

---

# Basics (no package)

```{r base_func}
input <- read.csv("data/main.csv")
input = read.csv("data/main.csv")
# BOTH ARE EQUAL

colnames(input)

input[1,8] # row = 1, column = 8
```

---

```{r base_func2}
input[1:5, "product_id"] # row = 1 to 5, column named "product_id"

input[c(1,7),c("product_id", "period_id")] # row = 1 & 7, columns named "product_id" & "period_id"

# keeping one of the arguments empty ([,a] OR [a,]) gives you all rows
# and columns specified by a, or all columns and rows specified by a
input[5,] # row = 5, column = ALL

input$period_id[5] # choose column named period_id, 5th value (=row) only

```

---

# READING FLATFILES

```{r read, message=FALSE, warning=FALSE}
# BASE
raw_b = read.csv("data/main.csv")
# TIDYVERSE
raw_t = readr::read_csv("data/main.csv", col_types = cols())
# DATA.TABLE
raw_d = data.table::fread("data/main.csv")
```

---

# FILTER & SELECT
```{r filter}
# BASE
raw_b[raw_b$X1.sales < 551,"product_id"]

# TIDYVERSE
raw_t %>% 
    filter(X1.sales < 551) %>% 
    select(product_id)

# DATA.TABLE
raw_d[X1.sales < 551, product_id]
```

---

# ARRANGE
```{r arrange}
# BASE
raw_b = raw_b[order(raw_b$X1.sales),]
raw_b[1:5, "X1.sales"]

# TIDYVERSE
raw_t = raw_t %>% 
    arrange(X1.sales)
raw_t$X1.sales[1:5]

# DATA.TABLE
raw_d = raw_d[order(X1.sales)]
raw_d[1:5, X1.sales]

```

---

# MUTATE
```{r mutate}
# BASE
raw_b$diff = raw_b$X1.sales - raw_b$X2.sales
raw_b$diff[1:5]

# TIDYVERSE
raw_t = raw_t %>% 
    mutate(diff = X1.sales - X2.sales)
raw_t$diff[1:5]

# DATA.TABLE
raw_d[,diff := X1.sales - X2.sales][1:5,diff]

```

---

# GROUP BY & SUMMARISE
```{r aggregate}
# BASE
aggregate(X1.sales ~ cat, raw_b, sum)

# TIDYVERSE
raw_b %>% 
    group_by(cat) %>% 
    summarise(sales = sum(X1.sales))
```

---

```{r aggregate2}
# DATA.TABLE
raw_d[,.(sales = sum(X1.sales)), by = cat]

```

---

# SPREAD & GATHER

```{r gather}
# BASE = never heard of one
# TIDYVERSE gather
raw_ts = 
    raw_t %>%
    select(period_id, product_id, X1.sales, X2.sales) %>% 
    gather(key = "type", value = "sales", X1.sales, X2.sales)
raw_ts

```

---

```{r spread}
# TIDYVERSE spread
raw_ts %>% 
    spread("type", "sales")

```

---

```{r melt}
# DATA.TABLE "spread"
raw_dm =
    raw_d[,.(period_id, product_id, X1.sales, X2.sales)] %>% 
    melt(id.vars = c("period_id", "product_id"),
         measure.vars = c("X1.sales", "X2.sales"))
raw_dm

```

---

```{r dcast}
# DATA.TABLE "gather"
raw_dm %>% 
    dcast(period_id + product_id ~ variable)

```

---

# WRITE FLATFILES

```{r write, eval=FALSE}
# BASE
write.csv(raw_b, "data/main.csv")
# TIDYVERSE
readr::write_csv(raw_t, "data/main.csv")
# DATA.TABLE
data.table::fwrite(raw_d, "data/main.csv")
```

---

# Key words - multi-table and looping
- bind
- joins
- vectorisation

---

# BINDING
- implemented in base R
- just glue tables together

```{r prepare, include = FALSE}
tab1 = data.frame(a = 1:4, b = 5:8)
tab2 = data.frame(d = 2:5, g = c("a","b","c","d"))
tab3 = data.frame(a = 10:15, b = 20:25)
```


.pull-left[
```{r tab1}
    tab1
```
]

.pull-right[
```{r tab2}
    tab2
```
]
```{r tab3}
tab3
```

---

```{r binds}
rbind(tab1, tab3) # bind rows (need same column names)
cbind(tab1, tab2) # bind columns (need same number of rows)
```

---

# <a href="http://www.dofactory.com/sql/join">JOINS</a>
- think vlookup (svyhledat)
- joinig two tables
    - on **ID!**

<img src=www/joins.png>

---

```{r prepare_join, include = FALSE}
ref = fread("data/ref_table.csv")
```

```{r joins}
# BASE, control by all.x = T/F & all.y = T/F
joint_b = merge(raw_b, ref, by = "period_id")
c(original = ncol(raw_b), merged = ncol(joint_b))

# TIDYVERSE, control by different function names (left_join, ...)
joint_t = inner_join(raw_t, ref, by = "period_id")
c(original = ncol(raw_t), merged = ncol(joint_t))

# DATA.TABLE, control by nomatch, or different positions (X[Y] OR Y[x])
joint_d = raw_d[ref, on = "period_id"]
c(original = ncol(raw_d), merged = ncol(joint_d))

```

---

# Vectorisation
- avoid for loops!

<img src= www/purrr.jpg height = 450>

---

# Vectorisation
```{r vector}
file_names = list.files("data/sub", full.names = T)

# BASE
raw_b = lapply(file_names, read.csv) # list of tables
raw_b = do.call(rbind, raw_b) # rbind tables into one

# TIDYVERSE
raw_t = map_df(file_names, read.csv) # nice and simple

# DATA.TABLE
raw_d = as.data.table(file_names)[,fread(file_names), by = file_names]

```